An Empirical Analysis of the Role of Energy in Economic Growth

### 2014-12-30 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Edited the implications section (thoroughly).

Some developed economies (US) benefit little from including energy as a factor of production, 
while some others (UK) benefit in a minor way. 
For other developed economies (JP), the results of CD and CES models point in opposite directions.  
For NICs, the results are mixed; 
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa (ZA) due to lack of precision in determining $\gamma$.  
In one OPEC country (SA), adding energy clearly helps to explain economic growth, 
while in another (IR), the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (TZ) or strong indication otherwise (ZM).

@@

For most economies (UK, JP, CN, ZA, SA, IR, and TZ), 
Figure~\ref{fig:CES_ternary_graph} shows that the relative importance of 
production factors capital~($k$), labour~($l$), and energy~($e$) 
is different for the three different CES nestings (Equations~\ref{eq:CESkle}--\ref{eq:CESekl}). 
In other words, you cannot view two of the ternary plots 
in the rightmost three columns of Figure~\ref{fig:CES_ternary_graph} and predict the third. 
Nesting structure matters little for the economies of the United States and Zambia.

@@

**** If we stick with these tables, do we want to use $\sim\sim\sim$ to indicate uncertainty? ****

@@

**** I'm fine if we want to change to another form of these tables.
But, I found the presentation below to be a good way to think about the results.
My idea was to say ``hang on! We have several ways to get at some of these elasticities.
Let's compare all the different ways to look at the elasticity between any two factors of production.
Do they agree?  
Are they in conflict?''  
Another point to make is that these comparisons need to be done honestly. 
We should not claim anything when the data are uncertain.

Regarding Martin's question about space, 
I've seen other papers in \emph{Energy Economics} that are 51 printed journal pages. \citep{Saunders:2008aa}
I expect we are fine on length. --MKH ****

@@

**** Shall we have a section here discussing which function is ``best''? --MKH ****
Here is a first cut at some criteria and a recommendation.

Allows assymetric uncertainty estimation via resampling: all work fine.

Best match to historical data: CES. But CD is close. LinEx struggles.

No ambiguity from model structure: CD and LinEx OK here. 
CES has the undesireable characteristic that form (nesting) affects outcome.

Easy to estimate: CD and LinEx can be fitted with linear function. 
CES requires complex and time-consuming optimzation techniques.

Small uncertainties in fitted parameters: CD and LinEx better than CES.

Easy to respect constraints during fitting process: CD and CES easy. LinEx difficult.

Our assessment: None of the production functions is perfect. 
Probably CD is best compromise among ease of estimating, good (but not best) fit to historical data, and
small uncertainties in fitted parameters.

If precise fitting to historical data is paramount (and you're unconcerned about uncertainties
in fitted coefficients), CES is the best.

If you need to represent rebound effects (not just backfire), 
CES must be employed, 
because the CES function was found by \citet{Saunders:2008aa} to be the only one of the production
functions studied here that is capable of representing rebound rebound in addition to backfire.
However, its ability to do so is tied to the values of the elasticity of substitution.
Unfortunately, we find that value of the elasticity of substitution is often highly uncertain.
Thus, whether the CES function is helpful for estimating both rebound and backfire effects
is questionable.

@@

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LinEx) 
and non-linear optimization algorithms (CES), 
the resampling/bootstrapping analysis performed herein 
provides insight into the question of how accurately parameters are known.

@@

A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2--0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} ~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****

@@

@@

**** Do we want to summarize the discussion here? --MKH ****

>>>>

Some developed economies (United States) benefit little from including energy as a factor of production, 
while some others (United Kingdom) benefit in a minor way. 
For other developed economies (Japan), the results of CD and CES models point in opposite directions.  
For NICs, the results are mixed; 
China does not convincingly benefit from energy-augmented production functions, 
and results are uncertain for South Africa due to lack of precision in determining $\gamma$.  
In one OPEC country, Saudi Arabia, adding energy clearly helps to explain economic growth, 
while in another, Iran, the benefit of energy augmentation is small and uncertain. 
For LDCs there is a either a strong indication that energy-augmented production functions 
are beneficial as an explanation for growth (Tanzania) or strong indication otherwise (Zambia).

@@

For most economies (United Kingdom, Japan, China, South Africa, Saudi Arabia, Iran, and Tanzania), 
Figure~\ref{fig:CES_ternary_graph} shows that the relative importance of 
production factors capital~($k$), labour~($l$), and energy~($e$) 
is different for the three different CES nestings (Equations~\ref{eq:CESkle}--\ref{eq:CESekl}). 
In other words, you cannot view two of the ternary plots 
in the rightmost three columns of Figure~\ref{fig:CES_ternary_graph} and predict the third. 
Nesting structure matters little for the economies of the United States and Zambia.

@@

**** If we stick with these tables, do we want to use $\sim\sim\sim$ to indicate uncertainty? --MKH ****

@@

**** I'm fine if we want to change to another form of these tables.
But, I found the presentation below to be a good way to think about the results.
My idea was to say ``hang on! We have several ways to get at some of these elasticities.
Let's compare all the different ways to look at the elasticity between any two factors of production.
Do they agree?  
Are they in conflict?''  
Another point to make is that these comparisons need to be done honestly. 
We should not claim anything when the data are uncertain. --MKH ****

**** Regarding Martin's question about space, 
I've seen other papers in \emph{Energy Economics} that are 51 printed journal pages. \citep{Saunders:2008aa}
I expect we are fine on length. --MKH ****

@@

@@

In addition to the usual parameter estimation process 
by linear coefficient determination (Cobb-Douglas and LinEx) 
and non-linear optimization algorithms (CES), 
the resampling analysis performed herein 
provides insight into the questions of 
(a)~how accurately parameters are known and 
(b)~the implications of parameter imprecision.

@@

A significant result to emerge from this work is that large parameter uncertainty 
does not necessarily lead to poor model fit to historical economic output. 
To choose one example of many, 
the value of $\gamma$ for the Cobb-Douglas model for Tanzania (0.59) 
shows a resampling range of 0.2--0.9. 
(See Figure~\ref{fig:CD_Ternary_Scatter_Graph}.) 
Yet, Cobb-Douglas models for Tanzania, with and without energy, both 
(a)~fit historical data very well 
(white line compared to black line in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}) 
and 
(b)~exhibit a very narrow range of resampled fits 
(thin gray band in Tanzania row of Figure~\ref{fig:CD_GDP_Spaghetti_Graph}). 
This means that a good fit to historical data does not indicate 
that model parameters have small uncertainty. 
In fact, similar to Tanzania, 
many economies exhibit both large uncertainty in model parameters and
very good fit to historical data. 
(See Figures~\ref{fig:CD_lambda_theta_graph} ~\ref{fig:CD_Ternary_Scatter_Graph} 
compared to Figure~\ref{fig:CD_GDP_Spaghetti_Graph}
for the Cobb-Douglas models and 
Figure~\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_sigma_delta_graph}
compared to Figure~\ref{fig:CES_Spaghetti_Graph} for CES models.) 
This result is consistent across almost all economies and models. 
Thus, we conclude that great care must be taken when interpreting estimated parameters: 
it is very easy to claim more certainty than is warranted. 
Where large uncertainty is observed, 
additional work is needed to contextualize the results, 
for example, 
sector-based analysis or ****. **** Martin, add more here? ****

@@

Several criteria could be applied 
to the question of which production function is best
for describing the role of energy in economic growth:
best match to historical data, 
minimum ambiguity in model structure,
minimum uncertainty in model parameters, and
easy parameter estimation process.
In the sections below, we evaluate the Cobb-Douglas, CES, and LINEX
models using these criteria.

%------------------------------
\subsubsection{Best match to historical data} 
\label{sec:match_history}
%------------------------------

Figures~\ref{fig:CD_GDP_Spaghetti_Graph}, \ref{fig:CES_Spaghetti_Graph}, and
\ref{fig:LINEX_Spaghetti_Graph} indicate that the CES production function
is consistently better than Cobb-Douglas or LINEX in terms of matching historical data.
Cobb-Douglas is a close second, but
LINEX struggles, especially with the Saudi Arabian and Tanzanian economies.%
  \footnote{
  The order of these results is unsurprising: 
  CES contains 6 fitted parameters; Cobb-Douglas, 4; and LINEX, 3.
  We expect models with more parameters to provide a better fit to historical data.
  }$^,$%
  \footnote{
  We also note that in contrast to our approach, shared by \citet{Warr:2012cg},
  of fitting constant $a_0$ and $c_t$ values, 
  \citet{Kummel:2011aa} used logistic functions to generate time-varying
  $a_0$ and $c_t$ values.
  \citeauthor{Kummel:2011aa}'s approach will necessarily produce a better match to historical data,
  because the logistic curves involve several additional fitting parameters.
  }

Thus, CES is preferred on this criterion.

%------------------------------
\subsubsection{Minimum ambiguity in model structure and interpretation} 
\label{sec:min_ambiguity}
%------------------------------

For each production function, 
values of fitted parameters give rise to interpretations of macroeconomic dynamics
that, in turn, have significant implications for macroeconomic policy
and human well-being.
Thus, it is desirable that such interpretations have little ambiguity.

The energy-augmented CES production function presents ambiguity to the energy analyst
in the form of the choice of factor nesting.
We showed above that nesting can significantly affect 
the interpretation of macroecnomic dynamics, especially 
the relative importance of factors of production~(\ref{fig:CES_ternary_graph}) and 
the understanding of substitutibility 
among factors of production. (See Section~\ref{sec:elasticity_of_substitution}).
Thus, the CES production function has the undesireable characteristic 
that model structure (i.e., nesting) affects both parameter values 
and macroeconomic interpretation.
Neither the Cobb-Douglas nor the LINEX production function 
exhibits similar structural ambiguity.

Thus, Cobb-Douglas or LINEX are preferred on this criterion.

%------------------------------
\subsubsection{Minimum uncertainty in model parameters} 
\label{sec:min_uncertainty}
%------------------------------

Uncertainties in fitted parameters give rise to uncertainty 
in macroeconomic dynamics inferred from fitted parameters.
Thus, small uncertainty in fitted parameters is desirable. 
Figures~\ref{fig:CD_lambda_theta_graph}, \ref{fig:CD_Ternary_Scatter_Graph},
\ref{fig:CES_lambda_gamma_graph}--\ref{fig:CES_ternary_graph}, and
\ref{fig:LINEXq_Resample_Graph} show that 
Cobb-Douglas and LINEX have, by and large, 
smaller uncertainties in fitted parameters than the CES production function.
Thus, Cobb-Douglas and LINEX are preferred on this criterion.

%------------------------------
\subsubsection{Robust parameter estimation process} 
\label{sec:robust_estimation}
%------------------------------

It is desirable that the combination of production function and parameter estimation process
has the following characteristics:
affords easy mathematical and numerical implementation, 
provides an easy way to respect fitted parameter constraints, and
allows asymmetric estimates of fitted parameter uncertainty.

%..............................
\paragraph{Easy mathematical and numerical implementation} 
\label{sec:easy_implementation}
%..............................

For the Cobb-Douglas and LINEX models, 
logarithmic transformation of the production functions 
allows simple linear estimation of model parameters.
In contrast, it is impossible to fit the CES model linearly,
so parameters must be estimated by a more-complicated and time-consuming optimization process.
(See Section~\ref{sec:CES_Parameter_Estimation}.)
On the 40-node \texttt{dahl} supercomputer at Calvin College~\citep{Adams:2008aa},
complete analysis of all data shown in this paper
(including 1000 resamples for every model, every economy, and every CES nesting)
takes about 15 hours.
Approximately 98\% of the processing effort is spent on the CES models.%
  \footnote{
  All Cobb-Douglas and LINEX models run on a single node (using 4 processors per node) 
  in about 1.5 hours (6 processor-hours).
  All CES models run on 9 nodes (using 3 processors per node for a total of 27 processors) 
  in about 15 hours (405 processor-hours).
  }
Thus, Cobb-Douglas and LINEX are preferred on this criterion.

%..............................
\paragraph{Easy to respect parameter constraints} 
\label{sec:easy_constraints}
%..............................

It is relatively easy to implement constraints on fitting parameter 
for the Cobb-Douglas production function. (See Section~\ref{sec:CD_Parameter_Estimation}.)
For the CES prodcution function, 
the capability to respect parameter constraints is a feature of the 
\texttt{cesEst} function~\citep{Henningsen:2011td} and is, therefore,
relatively easy to implement.
In contrast, because output elasticities $\alpha$, $\beta$, and $\gamma$ 
are not directly available as parameters in the LINEX model
(see Equations~\ref{eq:LINEX} and~\ref{eq:LINEX_transformed}),
it is very difficult to enforce constraints during the fitting process.%
  \footnote{
  See \citet{Kummel:2011aa} for one approach to respecting parameter constraints 
  for the LINEX function. 
  }
Indeed, we found in Section~\ref{sec:LINEX_Parameter_Estimation}
that the LINEX model regularly violates parameter constraints.

Thus, Cobb-Douglas and CES are preferred on this criterion.

%..............................
\paragraph{Easy asymmetric uncertainties} 
\label{sec:asymmetric_uncertainty}
%..............................

As discussed in Section~\ref{sec:implications},
constraints on fitting parameters make it 
desirable that the combination of production function
and the parameter estimation process
allows estimation of asymmetric parameter uncertainties.
In the present work, we found that the resampling techniques described 
in Section~\ref{sec:Resampling} afford estimates 
of asymmetric parameter uncertainties. 
Thus, none of the prodcution functions are preferred over the others on this criterion.

@@

The evidence above indicates that 
there is no quick or easy answer to the question
``which production function is best for augmentation by energy?''
We find that the answer depends upon the objective of the analysis.
In our opinion, 
the Cobb-Douglas models provides a good compromise among the competing criteria of
ease of parameter estimation, good fit to historical data, and
small uncertainties in fitted parameters.

If precise fitting to historical data is paramount and uncertainty
in fitted coefficients, especially elasticities of substitution, is not an issue, 
the CES model is the best.

If you need to represent rebound effects (not just backfire), 
CES \emph{must} be employed, 
because the CES function was found by \citet{Saunders:2008aa} to be the only production
function studied here that is capable of representing rebound in addition to backfire.
However, the ability of CES to represent rebound effects 
is tied to the presence of elasticities of substitution ($\sigma$ and $\sigma_1$) in the CES model.
Unfortunately, we resampling analyses show that estimated values 
of elasticity of substitution are often highly uncertain.
Thus, whether the CES function is helpful for estimating rebound in addition to backfire effects
remains an open question.

The LINEX model is intriguing, because of its thermodynamic roots, 
but our implementation suffers from 
poor reproduction of historical data and
difficulty in respecting parameter constraints.
If the LINEX function is to be used, we recommend that 
\citeapos{Kummel:2011aa} fitting approach (logistic curves for $a_0$ and $c_t$) be employed
and that the results be thoroughly tested and compared against other production functions.
